Exploring the Role of Small Molecules in Biological Systems Using Network Approaches

### 2014-12-30 paper.tex
Fleshed out section on scaffold networks

The notion of a scaffold is common place in medicinal chemistry and is
usually considered to be a ring-containing substructure pruned of
sidechains that represents the core structure of a series of
molecules. An example is the tricyclic core common to a number of
antidepressants (e.g., amitriptylene and imipramine) shown in Figure
\ref{fig:tca-core}. Scaffolds allow one to summarize a collection of
molecules (such as in patent claims) and one can usually assume that
compounds with the same scaffold will share a common synthetic pathway
or even a common mechanism of action.

While a number of approaches have been described to generate scaffolds
\cite{Lewell:1998aa,Bemis:1996aa,Katritzky:2000wf}, a few approaches
have considered hierarchical decomposition of scaffolds. Examples
include the HierS method \cite{Wilkens:2005il} and the Scaffold Tree
method \cite{Schuffenhauer:2007oz}.

\ctable[cap={}, caption={An example of a scaffold. In this case, this
  scaffold is common to many tricyclic antidepressants}, figure, botcap,
label={fig:tca-core}]    
{c}
{}
{\includegraphics{img/tca-core}}

@@

\begin{itemize}
\item Networks on the chemistry side tend to be more for visualization
\item In many cases the network does not necessarily lead to nove output
\item A key utility of network approaches is the ability to integrate
  different data types such as small molecules and proteins
\item Networks represent an approach to representing multiple
  relationships in a single contruct - multiple properties of a
  relatinship between two entities can be encoded by edge properties
  (width, color, length). Similarly multiple properties of a given
  entity can be encoded in node properties (size, shape, color) 
\end{itemize}
\bibliography{paper}

>>>>

The notion of a scaffold is common place in medicinal chemistry and is
usually considered to be a ring-containing substructure pruned of
sidechains that represents the core structure of a series of
molecules. An example is the tricyclic core common to a number of
antidepressants (e.g., amitriptylene and imipramine) shown in Figure
\ref{fig:tca-core}. Scaffolds allow one to summarize a collection of
molecules (such as in patent claims) and one can usually assume that
compounds with the same scaffold will share a common synthetic pathway
or even a common mechanism of action.
\ctable[cap={}, caption={An example of a scaffold. In this case, this
  scaffold is common to many tricyclic antidepressants}, figure, botcap,
label={fig:tca-core}]    
{c}
{}
{\includegraphics{img/tca-core}}

While a number of approaches have been described to generate scaffolds
\cite{Lewell:1998aa,Bemis:1996aa,Katritzky:2000wf}, a few approaches
have considered hierarchical decomposition of scaffolds. Examples
include the HierS method \cite{Wilkens:2005il} and the Scaffold Tree
method \cite{Schuffenhauer:2007oz}. The latter generates a series of
scaffolds based on an iterative removal of rings. The output of this
method is a series of scaffolds arranged in a tree-like structure
whose root node represents the simplest ring obtainable from the
starting scaffold.  Two scaffolds are connected if there exists a
parent--child relationship between the two. Scaffold trees generated
by this method lend themselves naturally to network
visualizations. Scaffold Hunter \cite{Wetzel:2009uq} is an example of
a tool that employs a network representation of scaffold trees. Figure
\ref{fig:scafftree} shows an example of a scaffold tree generated from
a set of 602 pyruvate kinase inhibitors. The graph is colored based on
$\log AC_{50}$ and the data is obtained from Pubchem AID 361. This
visualization lends itself naturally to adding overlays such as
molecular properties either on leaf nodes or else as aggregated values
on non-leaf nodes. The scaffold tree has been applied in a number of
studies \cite{Wetzel:2009uq,Renner:2009wm} and the structure of the
tree can be used as a guide to exploring chemical space, focusing on
regions (as represented by a scaffold) where there is minimal activity
(a ``biological hole'') or high activity.

\ctable[cap={}, caption={An example of a scaffold tree displayed as a
  radial network, generated from a set of pyruvate kinase inhibitors.}, figure, botcap,
label={fig:scafftree}]    
{c}
{}
{\includegraphics[width=\linewidth]{img/scafftree-full}}

Varin et al \cite{Varin:2011ve} describe ``scaffold networks'' which
are similar in nature to scaffold trees but instead of a tree
structure where each scaffold has a single parent, a DAG is
constructed by considering all scaffolds at a given level of the
hierarchy. An example of a scaffold network generated from Alosetron
is shown in Figure \ref{fig:scaffnet}, where the blue structures are
those that would be generated using the Scaffold Tree algorithm and
the green structures would be the extra scaffolds generated using the
Scaffold Network algorithm. Using this approach, together with the
compound set enrichment method \cite{Varin:2010zh} the authors were
able to efficiently identify active scaffold and active molecules

\ctable[cap={}, caption={An example of the scaffolds generated using
  the Scaffold Tree and Scaffold Network algorithms, starting from
  Alosetron. Blue scaffolds are generated by both methods and green
  scaffolds are generated only by the Scaffold Network
  method. Modified from Figure 1 in Varin et al \cite{Varin:2011ve}.}, figure, botcap,
label={fig:scaffnet}]    
{c}
{}
{\includegraphics[width=0.75\linewidth]{img/scaffold-network}}

@@

\begin{itemize}
\item Networks on the chemistry side tend to be more for visualization
\item In many cases the network does not necessarily lead to nove output
\item A key utility of network approaches is the ability to integrate
  different data types such as small molecules and proteins
\item Networks represent an approach to representing multiple
  relationships in a single construct - multiple properties of a
  relationship between two entities can be encoded by edge properties
  (width, color, length). Similarly multiple properties of a given
  entity can be encoded in node properties (size, shape, color) 
\end{itemize}
\bibliography{paper}
